---
title: "Analysis pipeline for my 4th year (MBiol) project"
format:
  html:
    toc: true
    toc-location: body
editor: visual
execute:
  enabled: false
---

This quarto document contains a record of python scripts and individual commands run in the terminal (Bash/shell environment) required for data analysis for my 4th Year Project. I have annotated scripts for clarity, and where necessary I have used text to further outline methodological steps.

## (1) Extracting mt .fasta sequences from the European Nucleotide Archive (ENA)

To obtain sequence data from relevant publications, my initial approach was to directly contact the corresponding author and request access to the consensus mtDNA .fasta sequence generated in the study. Where I received no response from the corresponding author, in most cases I was able to manually extract the mt .fasta sequence by downloading the .bam file stored on ENA. Below is the code I used in my pipeline for downloading a .bam file and forming a consensus sequence from only the mtDNA region.

```{Bash}

#view how the mtDNA is stored/named (usually 'MT')
samtools view -H input.bam | grep '^@SQ'

#index the .bam file
samtools index yourfile.bam

#extract the mtDNA region from the .bam file
samtools view -b yourfile.bam MT > yourfile_MT.bam

#index the extracted mtDNA section of the .bam file
samtools index yourfile_MT.bam

#run 'consensus_calling' script to generate a consensus file 
PATH="$HOME/micromamba/envs/samtools_new/bin:$PATH" ./consensus_calling_humans.sh SAMPLE_NAME_NO_BAM



```

### (1.1) Consensus_calling script:

```{python}


SAMPLE=$1
LENGTH=16370

# Print sample name
sample="${SAMPLE}.bam"

# Calculate depth of coverage (mitochondrial genome)
depth=$(samtools depth "${SAMPLE}.bam" | awk -v L="$LENGTH" '{sum+=$3}END{print sum/L}')

# Calculate breadth of coverage (mitochondrial genome)
breadth=$(samtools depth "${SAMPLE}.bam" | wc -l | awk -v L="$LENGTH" '{print ($1/L)*100 "%"}')

echo "$sample $depth $breadth" >> coverage_statistics.txt

# Create consensus FASTA
samtools consensus -f FASTA -a -d 1 -c 0.66 --min-MQ 30 "${SAMPLE}.bam" -o "${SAMPLE}.fasta"

# Change FASTA header
sed "s/chrM/${SAMPLE}/g" "${SAMPLE}.fasta" > "${SAMPLE}_consensus.fasta"


```

To extract multiple .bam files simultaneously, I ran a 'for' loop, where for every \*.bam the script was run. As before, for humans the script is identical except 'LENGTH' is changed.

## (2) Calculating sequence_completeness for each sample

My mtDNA sequences were obtained in a .fasta file format from a variety of sources and publications. Methods of calculating coverage_depth vary between publications, therefore I required a method for estimating coverage_depth that would allow for a standardised comparison across samples. Additionally, for a large number of samples no information on coverage_depth or coverage_breadth at the single-sequence level was reported. Since in most cases I did not have access to the .bam file, I was unable to calculate coverage_depth or coverage_breadth de-novo. Therefore, I devised a method of calculating sequence_completeness from only the mt .fasta file, which could act as a proxy for coverage_depth that could be fairly compared across samples. The code and method is identical for humans, except genome length differs.

The script calculates sequence_completeness for each sample, which is the proportion of nucleotide bases (ATCG) in relation to the expected mitochondrial genome length. This script treats both ambiguous bases (Ns) and gaps (-s) as missing data, which do not count towards coverage.

```{python}

#Calculate coverage_depth: method using number of non-N bases/expected mt genome length

import sys

DOG_MT_LENGTH = 16727  # dog mitochondrial genome length (humans would require 16569)

def fasta_records(path):
    h, s = None, []
    with open(path) as f:
        for line in f:
            line = line.strip()
            if not line:
                continue
            if line.startswith(">"):
                if h:
                    yield h, "".join(s)
                h = line[1:]
                s = []
            else:
                s.append(line)
        if h:
            yield h, "".join(s)

if len(sys.argv) != 2:
    sys.exit("usage: calc_dog_mt_completeness.py <fasta>")

for header, seq in fasta_records(sys.argv[1]):
    acgt = sum(1 for b in seq.upper() if b in "ACGT")
    completeness = acgt / DOG_MT_LENGTH
    print(header + "\t" + "{:.4f}".format(completeness))




```

## (3)Sequence Alignment on MAFFT

I performed the Initial sequence alignment using all samples. After a series of filtering steps, I repeated the alignment in subsequent iterations, removing low coverage samples. I used the following line of code for all alignments:

```{bash}

mafft --auto --thread 20 $FASTA
```

## (4)Phylogenetic tree generation

### (4.1) ML phylogeny generation using IQ-TREE (v2.2.6)

I used this code to generate a 1000 bootstrap replicate phylogeny. I used the coyote mitogenome (NC_008093.1) as an outgroup, which I downloaded from NCBI.

```{bash}

~/bin/iqtree-2.2.2.6-Linux/bin/iqtree2 -s ~/dogs/alignment/alignment3/alignment_dogs_172_nodups.fasta -st DNA -o NC_008093.1 -B 1000 -ntmax 20 -T AUTO -m TEST &


```

I downloaded consensus trees in the Newick format and initially visualised them using FigTree (v1.4.4).

### (4.2) BEAST phylogeny generation

I used the following code to generate BEAST phylogenetic trees. I generated .xml files using the BEAUti (v1.10.4) graphical interface.

```{bash}

/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/beast -threads 20 aligned_file.xml
```

#### Dogs

I ran three independent iterations of code. I combined the .log output files using logcombiner (code below). I inspected the ESS values of the resulting combined.log file. The critical threshold value of 200 was exceeded by all parameters, indicating that we could be confident in moving forward with the outputted BEAST tree in later stages of our analysis.

```{bash}

#Merge log files
/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/logcombiner \
-burnin 10 \
dogs_iteration4_run1.log \
dogs_iteration4_run2.log \
dogs_iteration4_run3.log \
dogs_iteration4_combined.log
```

Logcombiner (v1.10.4) was used combine the .trees files from the three independent runs:

```{bash}

#Merge tree files
/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/logcombiner \
-trees \
-burnin 10 \
dogs_iteration4_run1.trees \
dogs_iteration4_run2.trees \
dogs_iteration4_run3.trees \
dogs_iteration4_combined.trees
```

TreeAnnotator (v1.10.4) was used to extract the Maximum Clade Credibility (MCC) summary tree for the distribution:

```{bash}

#Create consensus tree file - MCC from combined runs 
/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/treeannotator \
-burnin 10 \
-heights median \
dogs_iteration4_combined.trees \
dogs_iteration4_MCC.tree
```

#### Humans

Initially I ran three independent iterations of code, however due to the higher number of human samples, we decided to include an additional run at a 60-million-chain length so there were four .trees files to combine using Logcombiner.

```{bash}

#Merge log files
/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/logcombiner \
-burnin 10 \
humans_it14_517_beast_20M_run1.log \
humans_it14_517_beast_20M_run2.log \
humans_it14_517_20M_run3.log \
humans_it14_517_60M_run1.log \
humans_it14_517_3_20M_1_60M_combined.log
```

```{bash}

#Merge tree files
/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/logcombiner \
-trees \
-burnin 10 \
humans_it14_517_beast_20M_run1.trees \
humans_it14_517_beast_20M_run2.trees \
humans_it14_517_20M_run3.trees \
humans_it14_517_60M_run1.trees \
humans_it14_517_3_20M_1_60M_combined.trees
```

The increased chain length meant additional steps to thin the .trees files were needed before an MCC summary tree could be extracted, which was achieved using the code below:

```{bash}

/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/logcombiner -trees -burnin 0 -resample 5000 \
  humans_it14_517_3_20M_1_60M_combined.trees \
  humans_it14_517_3_20M_1_60M_thinned.trees

/media/jbod2/eugenio/local_bin/BEASTv1.10.4/bin/treeannotator \
  -burninTrees 0 \
  -heights median \
  humans_it14_517_3_20M_1_60M_thinned.trees \
  humans_it14_517_allruns_MCC.tree
```

# (5) Statistical Analysis

## (5.1) Generating Genetic Distance Matrices

### Dogs

According to the generated .iqtreefile, the best fit model according to the Bayesian Information Criterion (BIC) for dogs was HKY+F+G4. I used the Maximum Likelihood (ML) *.tree* file to generate a model-based genetic distance matrix consistent with the ML phylogeny under the HKY+F+G4 model. I used the following code, which generates a patristic distance matrix, in which each value represents the sum of branch lengths connecting any two tips.

Two samples were removed from the final genetic distance matrix:

1.  the coyote outgroup (NC_008093.1)

2.  one sample from Harbin, China (China_Harbin_12904_Heilongjiang_China_12904_A2_A2c),

The Chinese sample was included in the phylogeny to act like an outgroup, allowing us to analyse how Pre-contact dog samples would cluster with a more basal Eurasian Sample. However, it was excluded from statistical analysis because we aimed to focus only on relationships between humans and dogs restricted to within North and South America.

```{r}
# make_dist_matrix.R
# Computes pairwise patristic distances from an ML tree and writes CSV.


library(ape)

# change filename if needed
ml_tree_file <- "dogs_iteration5_ML.tree"
out_csv <- "dogs_iteration5_genetic_distance_matrix.csv"

# read tree
tr <- read.tree(ml_tree_file)

# compute pairwise patristic distances (in substitutions/site)
dist_mat <- cophenetic.phylo(tr)

# ensure row/col names are unchanged
rownames(dist_mat) <- tr$tip.label
colnames(dist_mat) <- tr$tip.label

# write CSV (row names preserved)
write.csv(dist_mat, out_csv, quote = FALSE)
cat("Wrote pairwise distance matrix to", out_csv, "\n")


#write new csv which filters out outgroup and chinese sample

# Read distance matrix
dist_mat <- read.csv("dogs_iteration5_genetic_distance_matrix.csv", 
                     row.names = 1, 
                     check.names = FALSE)

# Samples to remove
remove_samples <- c(
  "NC_008093.1_extraction",
  "China_Harbin_12904_Heilongjiang_China_12904_A2_A2c"
)

# Remove rows and columns
dist_mat_filtered <- dist_mat[!(rownames(dist_mat) %in% remove_samples),
                              !(colnames(dist_mat) %in% remove_samples)]

# Write new file
write.csv(dist_mat_filtered,
          "dogs_iteration5_genetic_distance_matrix_filtered.csv",
          quote = FALSE)



#write new csv to format ids without _extraction

# Read distance matrix
dist_mat <- read.csv("dogs_iteration5_genetic_distance_matrix_filtered.csv",
                     row.names = 1,
                     check.names = FALSE)

# Remove "_extraction" from row and column names
rownames(dist_mat) <- gsub("_extraction", "", rownames(dist_mat))
colnames(dist_mat) <- gsub("_extraction", "", colnames(dist_mat))

# Write updated file
write.csv(dist_mat,
          "dogs_iteration5_genetic_distance_matrix_filtered_clean.csv",
          quote = FALSE)
```

### Humans

For humans, I generated a genetic distance matrix from an ML tree inferred under the same model (HKY+F+G4) to maintain consistency with the dogs. The Yoruba outgroup (AF347014.1) was removed from the final genetic distance matrix.

```{r}
# make_dist_matrix_humans.R
# Computes pairwise patristic distances from an ML tree and writes CSV.

library(ape)

ml_tree_file <- "humans_iteration6_ML.tree"
out_csv <- "humans_iteration6_genetic_distance_matrix.csv"

# read tree
tr <- read.tree(ml_tree_file)

# compute pairwise patristic distances (in substitutions/site)
dist_mat <- cophenetic.phylo(tr)

# ensure row/col names are unchanged
rownames(dist_mat) <- tr$tip.label
colnames(dist_mat) <- tr$tip.label

# write CSV (row names preserved)
write.csv(dist_mat, out_csv, quote = FALSE)

cat("Wrote pairwise distance matrix to", out_csv, "\n")


#-------delete the outgroup

# Read human distance matrix
dist_mat <- read.csv("humans_iteration6_genetic_distance_matrix.csv",
                     row.names = 1,
                     check.names = FALSE)

# Sample to remove
remove_sample <- "AF347014.1_yoruba_reference"

# Remove row and column
dist_mat_filtered <- dist_mat[!(rownames(dist_mat) %in% remove_sample),
                              !(colnames(dist_mat) %in% remove_sample)]

# Write updated file
write.csv(dist_mat_filtered,
          "humans_iteration6_genetic_distance_matrix_filtered.csv",
          quote = FALSE)

cat("Removed", remove_sample, "and wrote filtered matrix.\n")


```
