---
title: "R scripts for phylogeny visualisation for my 4th year (MBiol) project"
format:
  html:
    self-contained: true 
    toc: true
    toc-location: body
    css: styles.css
editor: visual
execute:
  warning: false
---

This Quarto document contains a record of the R scripts used to produce the final phylogenies presented as figures in my 4th Year Project. The scripts have been annotated for clarity.

# Phylogenetic trees

Phylogenetic trees were initially inspected and rooted in FigTree (v1.4.4), however R code was used to build and annotate tree plots.

## Dogs

### ML tree

```{r}
#| results: 'hide'
#| fig-width: 11
#| fig-height: 30
#| fig-cap: "ML tree of dog samples, nodes annotated with Bootstrap scores (shown when >50)"
#| fig-align: center


library(ape)
library(dplyr)
library(readr)
library(stringr)
library(ggtree)
library(ggplot2)

# --- paths / params ---
tree_file   <- "dogs_it7_ML.tree"
meta_file   <- "dogs_it7_metadata.csv"
out_csv     <- "dogs_it7_clades_output.csv"
boot_thresh <- 71

# --- read inputs & clean IDs ---
tree <- read.tree(tree_file)
tree$tip.label <- gsub("_extraction$", "", tree$tip.label)
tree$tip.label <- trimws(tree$tip.label)

meta <- read_csv(meta_file, show_col_types = FALSE)
meta$Fasta_ID_alignment <- trimws(meta$Fasta_ID_alignment)
meta <- meta %>% rename(tip = Fasta_ID_alignment)
hg_col <- "Haplogroup"

# --- helper: descendant tip labels for a node ---
get_tips_for_node <- function(tr, node){
  Ntip <- length(tr$tip.label)
  children <- tr$edge[tr$edge[,1]==node,2]
  stack <- as.integer(children)
  tips <- character(0)
  while(length(stack)){
    v <- stack[1]; stack <- stack[-1]
    if(v <= Ntip) tips <- c(tips, tr$tip.label[v])
    else stack <- c(stack, tr$edge[tr$edge[,1]==v,2])
  }
  unique(tips)
}

# --- helper: all descendant nodes ---
get_all_desc_nodes <- function(tr, node){
  children <- tr$edge[tr$edge[,1]==node,2]
  stack <- as.integer(children)
  nodes <- integer(0)
  while(length(stack)){
    v <- stack[1]; stack <- stack[-1]
    nodes <- c(nodes, v)
    if(!(v <= length(tr$tip.label))) stack <- c(stack, tr$edge[tr$edge[,1]==v,2])
  }
  unique(nodes)
}

# --- hap lookup ---
hap_lookup <- setNames(meta[[hg_col]], meta$tip)

# --- identify qualifying clades ---
if(is.null(tree$node.label)) stop("No internal node labels found.")
Ntip <- length(tree$tip.label)
Nnode <- tree$Nnode
results <- vector("list", 0)

for(node in (Ntip+1):(Ntip+Nnode)){
  idx <- node - Ntip
  boot_raw <- tree$node.label[idx]
  boot <- suppressWarnings(as.numeric(boot_raw))
  if(is.na(boot)) next
  tips <- get_tips_for_node(tree, node)
  hgs <- hap_lookup[tips]
  all_same_hg <- length(na.omit(unique(hgs)))==1 && !any(is.na(hgs))
  hg_val <- if(all_same_hg) unique(hgs) else NA_character_
  qualifies <- (!is.na(boot) && boot >= boot_thresh && all_same_hg)
  results[[length(results)+1]] <- data.frame(
    node = node,
    bootstrap = boot,
    haplogroup = hg_val,
    n_tips = length(tips),
    tips = paste(tips, collapse=";"),
    qualifies = qualifies,
    stringsAsFactors = FALSE
  )
}

res_df <- bind_rows(results) %>% arrange(desc(bootstrap), node)
write_csv(res_df, out_csv)
cat("Wrote", out_csv, "\n")

# --- plotting ---
clades <- res_df %>% filter(qualifies == TRUE)

p_base <- ggtree(tree,
                 layout = "rectangular",
                 branch.length = "none",
                 size = 0)

td <- p_base$data
td$clade_hap <- NA_character_

if(nrow(clades) > 0){
  for(i in seq_len(nrow(clades))){
    node_i <- clades$node[i]
    hap_i  <- clades$haplogroup[i]
    desc_nodes <- c(node_i, get_all_desc_nodes(tree, node_i))
    td$clade_hap[td$node %in% desc_nodes] <- hap_i
  }
}

# --- manually override specific tips to black ---
override_tips <- c(
  "Az-563_TizayucaR5_Mexico_c_1100_A2b1a",
  "SG0892_L1_A1_ReedXI_USA_c_525",
  "D_California1_5000_NorthHead_USA_c_5000",
  "D_California4_4000_CanadaVerde_USA_c_4000",
  "D_JaguarCave1_1400_JaguarCave_USA_d_1400_A2b3",
  "D_Mississippi1_675_Oliver_USA_c_365",
  "D_AngelMound4_1000_AngelMounds_USA_c_1000_A2b",
  "D_Colorado1_1100_GrassMesa_USA_c_1100_A2b1b",
  "ELOLIV-S-R27_El_Olivar_Chile_c_850"
)

td$clade_hap[td$isTip & td$label %in% override_tips] <- NA
td$clade_hap[td$node == 277] <- NA
td$clade_hap[td$parent == 277] <- NA
td$clade_hap[td$node == 263] <- NA
td$clade_hap[td$node == 264] <- NA
td$clade_hap[td$node == 225] <- NA


p_base$data <- td

# --- palette ---
unique_haps <- na.omit(unique(td$clade_hap))

cb_palette <- c(
  "#E7180B",
  "#377CEF", 
  "#FFA1AD", 
  "#7B3B97",
  "#007C66",
  "#FFA600", 
  "#8A4F2A", 
  "#000000"  
)

cols <- setNames(cb_palette[1:length(unique_haps)], unique_haps)

p2 <- p_base +
  geom_tree(aes(color = clade_hap), linewidth = 0.4, show.legend = FALSE) +
  geom_tiplab(aes(label=label, color = clade_hap), size = 3) +
  geom_text2(aes(label = label, color = clade_hap),
             data = function(x) subset(x, !isTip & !is.na(label)),
             size = 2,
             hjust = -0.3) +
  scale_color_manual(values = cols, na.value = "black") +
  theme_tree() +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    legend.position = "none"
  ) + 
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9)))


p2


```

### BEAST tree

```{r}
#| fig-width: 11
#| fig-height: 30
#| fig-cap: "BEAST tree of dog samples, nodes annotated with posterior support values (shown when >0.5)"
#| fig-align: center

# Code to build a BEAST tree, colouring monophyletic clades using manually defined nodes 


# Load packages
library(treeio)
library(ggtree)
library(ggplot2)
library(readr)
library(dplyr)
library(ape)

# -------- helper function --------
getDescendants <- function(tree, node) {
  children <- tree$edge[tree$edge[,1] == node, 2]
  all_desc <- children
  for (child in children) {
    all_desc <- c(all_desc, getDescendants(tree, child))
  }
  return(all_desc)
}
# ---------------------------------

# Load BEAST MCC tree
beast_tree <- read.beast("dogs_it7_BEAST.nexus")
tree <- as.phylo(beast_tree)
beast_tree@phylo$tip.label <- gsub("_extraction$", "", beast_tree@phylo$tip.label)

# Load manual clade definitions (node, colour, hex_code)
clade_defs <- read_csv("dogs_it7_BEAST_manual_clade_definitions.csv",
                       show_col_types = FALSE)

# Initialise colour vector for all nodes
node_colours <- rep(NA, Ntip(tree) + Nnode(tree))

# Assign clade colours to each defined node + all descendants
for (i in seq_len(nrow(clade_defs))) {
  node_id <- clade_defs$node[i]
  hex     <- clade_defs$hex_code[i]
  desc <- getDescendants(tree, node_id)
  node_colours[node_id] <- hex
  node_colours[desc] <- hex
}

# Build base ggtree object and ensure plotting data has labels/posterior
p1 <- ggtree(beast_tree, layout = "rectangular", branch.length = "none")

# fortify/be safe: attach our clade_colour vector to the plotting dataframe
p1$data$clade_colour <- node_colours[p1$data$node]

# Ensure every row has a label (use beast_tree@data if needed)
if (! "label" %in% names(p1$data) || all(is.na(p1$data$label))) {
  p1$data$label <- beast_tree@data$label
} else {
  missing_labels <- is.na(p1$data$label) & !is.null(beast_tree@data$label)
  if (any(missing_labels)) p1$data$label[missing_labels] <- beast_tree@data$label[missing_labels]
}

# Ensure posterior column exists for internal nodes (fill from beast_tree if needed)
if (! "posterior" %in% names(p1$data)) p1$data$posterior <- NA
if (! is.null(beast_tree@data$posterior)) {
  missing_post <- is.na(p1$data$posterior) & !is.na(beast_tree@data$posterior)
  if (any(missing_post)) p1$data$posterior[missing_post] <- beast_tree@data$posterior[missing_post]
}

# Set default colour = black for any node/tip not assigned a clade colour
p1$data$clade_colour[is.na(p1$data$clade_colour)] <- "#000000"

# Now plot, using the clade_colour already stored in p$data
p1.5 <- p1 +
  geom_tree(aes(colour = clade_colour)) +
  geom_tiplab(aes(label = label, colour = clade_colour),
              size = 3,
              align = TRUE,
              linetype = "dotted",
              show.legend = FALSE) +
  scale_colour_identity() +
  geom_text2(aes(subset = !isTip,
                 label = ifelse(is.na(posterior), "", sprintf("%.2f", posterior)),
                 colour = clade_colour),
             size = 2,
             hjust = -0.3,
             na.rm = TRUE) +
  coord_cartesian(clip = "off") +
  theme_tree() +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9)))

p1.5
```

## Humans

### ML tree

```{r}
#| results: 'hide'
#| fig-width: 15
#| fig-height: 60
#| fig-cap: "ML tree of human samples, nodes annotated with Bootstrap scores (shown when >50)"
#| fig-align: center



library(ape)
library(ggtree)
library(ggplot2)
library(readr)
library(dplyr)

# -------- helper function --------
getDescendants <- function(tree, node) {
  children <- tree$edge[tree$edge[, 1] == node, 2]
  all_desc <- children
  for (child in children) {
    all_desc <- c(all_desc, getDescendants(tree, child))
  }
  return(all_desc)
}
# ---------------------------------

# Nodes to emphasise
highlight_nodes <- c(521, 663, 667, 858, 783, 805, 794, 785)

# Load ML tree
tree <- read.tree("humans_it14_517_aligned_inspected_ML.tree")

# Clean tip labels if needed
tree$tip.label <- gsub("_extraction$", "", tree$tip.label)
tree$tip.label <- trimws(tree$tip.label)
tree$tip.label <- gsub("\u2212", "-", tree$tip.label)

# Root tree with outgroup
tree <- root(tree, outgroup = "AF347014.1_yoruba_reference", resolve.root = TRUE)

# Remove any "Root" label
if (!is.null(tree$node.label)) {
  tree$node.label[tree$node.label == "Root"] <- ""
}

# Read tip colouring table
colour_df <- read_csv("humans_it14_colouring_ML_based_on_beast.csv", show_col_types = FALSE) %>%
  mutate(
    Fasta_ID_alignment = trimws(as.character(Fasta_ID_alignment)),
    Fasta_ID_alignment = gsub("\u2212", "-", Fasta_ID_alignment),
    Assigned_Hg_based_on_tree = na_if(trimws(as.character(Assigned_Hg_based_on_tree)), ""),
    colour_name = na_if(trimws(as.character(colour_name)), ""),
    colour_hex_code = na_if(trimws(as.character(colour_hex_code)), "")
  )

# Exact ID lookup for tip labels
colour_lookup <- setNames(colour_df$colour_hex_code, colour_df$Fasta_ID_alignment)

# Match colours to tree tips; default to black if no hex code or no match
tip_colours <- colour_lookup[tree$tip.label]
tip_colours[is.na(tip_colours) | tip_colours == ""] <- "#000000"

# Read internal-node colouring table
internal_df <- read_csv("colouring_internal_nodes_ML.csv", show_col_types = FALSE) %>%
  mutate(
    clade_defining_node_number = suppressWarnings(as.integer(clade_defining_node_number)),
    colour_name = na_if(trimws(as.character(colour_name)), ""),
    colour_hex_code = na_if(trimws(as.character(colour_hex_code)), "")
  ) %>%
  filter(!is.na(clade_defining_node_number), !is.na(colour_hex_code))

# Base plot
p4 <- ggtree(
  tree,
  layout = "rectangular",
  branch.length = "none",
  size = 0.4
)

# Attach tip colours to plotting data
p4$data$tip_colour <- tip_colours[p4$data$label]
p4$data$tip_colour[is.na(p4$data$tip_colour) | p4$data$tip_colour == ""] <- "#000000"

# Default all branches to black
p4$data$branch_colour <- "#000000"

# Colour all branches that fall within / stem from each internal node
# Later rows overwrite earlier ones if clades overlap
for (i in seq_len(nrow(internal_df))) {
  node_i <- internal_df$clade_defining_node_number[i]
  hex_i   <- internal_df$colour_hex_code[i]
  
  desc_nodes <- c(node_i, getDescendants(tree, node_i))
  p4$data$branch_colour[p4$data$node %in% desc_nodes] <- hex_i
}

# Add bootstrap values to plotting data
p4$data$bootstrap <- NA_real_
if (!is.null(tree$node.label)) {
  internal_nodes <- (Ntip(tree) + 1):(Ntip(tree) + tree$Nnode)
  p4$data$bootstrap[p4$data$node %in% internal_nodes] <- suppressWarnings(as.numeric(tree$node.label))
}

# Bootstrap scores >= 50 only
p4$data$show_bootstrap <- !p4$data$isTip & !is.na(p4$data$bootstrap) & p4$data$bootstrap >= 50
p4$data$bootstrap_label <- ifelse(p4$data$show_bootstrap, sprintf("%.0f", p4$data$bootstrap), "")

# Highlight specific nodes
p4$data$is_highlight <- p4$data$node %in% highlight_nodes

# Use the branch colour for the bootstrap text too
p4$data$bootstrap_colour <- p4$data$branch_colour

# Plot
p4.5 <- p4 +
  geom_tree(aes(colour = branch_colour)) +
  geom_tiplab(
    aes(label = label, colour = tip_colour),
    size = 3,
    align = TRUE,
    linetype = "dotted",
    show.legend = FALSE
  ) +
  # normal bootstrap labels
  geom_text2(
    aes(
      subset = show_bootstrap & !is_highlight,
      label = bootstrap_label,
      colour = bootstrap_colour
    ),
    size = 2,
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  # highlighted bootstrap labels
  geom_text2(
    aes(
      subset = show_bootstrap & is_highlight,
      label = bootstrap_label,
      colour = bootstrap_colour
    ),
    size = 5,
    fontface = "bold",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  scale_colour_identity() +
  coord_cartesian(clip = "off") +
  theme_tree() +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(20, 60, 20, 40)
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9)))

p4.5

```

### BEAST tree

```{r}
#| results: 'hide'
#| fig-width: 15
#| fig-height: 60
#| fig-cap: "BEAST tree of human samples, nodes annotated with posterior support values (shown when >0.5)"
#| fig-align: center


library(treeio)
library(ggtree)
library(ggplot2)
library(ape)
library(readr)
library(dplyr)

# -------- helper function --------
getDescendants <- function(tree, node) {
  children <- tree$edge[tree$edge[, 1] == node, 2]
  all_desc <- children
  for (child in children) {
    all_desc <- c(all_desc, getDescendants(tree, child))
  }
  return(all_desc)
}
# ---------------------------------

# Nodes to emphasise
highlight_nodes <- c(637, 521, 779, 784, 959, 981, 972, 960)

# Load BEAST MCC tree (already rooted)
beast_tree <- read.beast("humans_it14_517_allruns_MCC_rooted.tree")
tree <- as.phylo(beast_tree)

# Clean tip labels if needed
tree$tip.label <- gsub("_extraction$", "", tree$tip.label)
beast_tree@phylo$tip.label <- tree$tip.label

# Read clade colouring table
clade_defs <- read_csv("human_it14_beast_tree_colour.csv", show_col_types = FALSE) %>%
  mutate(
    Fasta_ID_alignment = as.character(Fasta_ID_alignment),
    Assigned_Hg_based_on_tree = na_if(trimws(as.character(Assigned_Hg_based_on_tree)), ""),
    node_of_assigned_monophyletic_clade = suppressWarnings(as.integer(node_of_assigned_monophyletic_clade)),
    colour_name = na_if(trimws(as.character(colour_name)), ""),
    colour_hex_code = na_if(trimws(as.character(colour_hex_code)), "")
  )

# Keep only rows with a valid clade node and colour
clade_rows <- clade_defs %>%
  filter(
    !is.na(Assigned_Hg_based_on_tree),
    !is.na(node_of_assigned_monophyletic_clade),
    !is.na(colour_hex_code)
  )

# Default all nodes/tips to black
node_colours <- rep("#000000", Ntip(tree) + Nnode(tree))

# Assign clade colours to each defined node + all descendants
for (i in seq_len(nrow(clade_rows))) {
  node_id <- clade_rows$node_of_assigned_monophyletic_clade[i]
  hex <- clade_rows$colour_hex_code[i]
  desc <- getDescendants(tree, node_id)
  
  node_colours[node_id] <- hex
  if (length(desc) > 0) node_colours[desc] <- hex
}

# Build base ggtree object
p3 <- ggtree(beast_tree, layout = "rectangular", branch.length = "none")

# Ensure plotting data has labels
if (!"label" %in% names(p3$data) || all(is.na(p3$data$label))) {
  p3$data$label <- beast_tree@data$label
} else {
  missing_labels <- is.na(p3$data$label) & !is.null(beast_tree@data$label)
  if (any(missing_labels)) p3$data$label[missing_labels] <- beast_tree@data$label[missing_labels]
}

# Ensure posterior column exists
if (!"posterior" %in% names(p3$data)) p3$data$posterior <- NA
if (!is.null(beast_tree@data$posterior)) {
  missing_post <- is.na(p3$data$posterior) & !is.na(beast_tree@data$posterior)
  if (any(missing_post)) p3$data$posterior[missing_post] <- beast_tree@data$posterior[missing_post]
}

# Attach clade colours
p3$data$clade_colour <- node_colours[p3$data$node]

# Show posterior >= 0.50
p3$data$show_support <- !p3$data$isTip & !is.na(p3$data$posterior) &
  (p3$data$posterior >= 0.5)

# Labels
p3$data$support_label <- ifelse(
  p3$data$show_support,
  sprintf("%.2f", p3$data$posterior),
  ""
)

# Identify highlighted nodes
p3$data$is_highlight <- p3$data$node %in% highlight_nodes

# Support colour matches clade colour
p3$data$support_colour <- p3$data$clade_colour

# Plot
p3.5 <- p3 +
  geom_tree(aes(colour = clade_colour)) +
  geom_tiplab(
    aes(label = label, colour = clade_colour),
    size = 3,
    align = TRUE,
    linetype = "dotted",
    show.legend = FALSE
  ) +
  # normal posterior labels
  geom_text2(
    aes(
      subset = show_support & !is_highlight,
      label = support_label,
      colour = support_colour
    ),
    size = 2,
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  # highlighted posterior labels
  geom_text2(
    aes(
      subset = show_support & is_highlight,
      label = support_label,
      colour = support_colour
    ),
    size = 5,
    fontface = "bold",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  scale_colour_identity() +
  coord_cartesian(clip = "off") +
  theme_tree() +
  theme(
    plot.margin = margin(20, 60, 20, 40)
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9)))

p3.5

```

# Comparing ML and BEAST trees

I generated a figure to assess how well the ML tree supports the BEAST tree topology. The BEAST phylogeny is shown as the base, and circles are plotted on key clade-defining nodes. A dark blue circle indicates that 95% of the BEAST clade is recreated in the ML phylogeny, so there is 95% similarity in the placement of taxa within that monophyletic clade. This gives us confidence that the BEAST-defined topology is recovered by the corresponding ML clades.

## Dogs

```{r}
#| fig-width: 11
#| fig-height: 30
#| fig-cap: "Tree comparing dog ML and BEAST tree topology. Foundation is the BEAST tree with circles on major nodes correspondign to ML agreement of taxa inside major clades (dark blue = 95% similarity, light blue = > 50% similarity, white = <50% similarity)."
#| fig-align: center

library(treeio)
library(ggtree)
library(ggplot2)
library(readr)
library(dplyr)
library(ape)

options(ignore.negative.edge = TRUE)

relaxed_thresh <- 0.95

# -------- helper functions ----------
get_descendant_tips <- function(phy, node) {
  children <- phy$edge[phy$edge[, 1] == node, 2]
  tips <- integer(0)
  for (ch in children) {
    if (ch <= length(phy$tip.label)) {
      tips <- c(tips, ch)
    } else {
      tips <- c(tips, get_descendant_tips(phy, ch))
    }
  }
  unique(tips)
}

get_descendant_labels <- function(phy, node) {
  idx <- get_descendant_tips(phy, node)
  phy$tip.label[idx]
}

getDescendants <- function(tree, node) {
  children <- tree$edge[tree$edge[, 1] == node, 2]
  all_desc <- children
  for (child in children) {
    all_desc <- c(all_desc, getDescendants(tree, child))
  }
  return(all_desc)
}
# ------------------------------------

# -------- load BEAST tree ----------
beast_tree <- read.beast("dogs_it7_BEAST.nexus")
tree <- as.phylo(beast_tree)

# Clean tip labels if needed
tree$tip.label <- gsub("_extraction$", "", tree$tip.label)
beast_tree@phylo$tip.label <- tree$tip.label

# -------- read dog BEAST colouring table ----------
clade_defs <- read_csv("dogs_it7_BEAST_manual_clade_definitions.csv", show_col_types = FALSE) %>%
  mutate(
    node = suppressWarnings(as.integer(node)),
    hex_code = na_if(trimws(as.character(hex_code)), "")
  )

clade_rows <- clade_defs %>%
  filter(!is.na(node), !is.na(hex_code))

# Default all nodes/tips to black
node_colours <- rep("#000000", Ntip(tree) + Nnode(tree))

# Assign clade colours to each defined node + all descendants
for (i in seq_len(nrow(clade_rows))) {
  node_id <- clade_rows$node[i]
  hex <- clade_rows$hex_code[i]
  desc <- getDescendants(tree, node_id)
  
  node_colours[node_id] <- hex
  if (length(desc) > 0) node_colours[desc] <- hex
}

# -------- read tanglegram table ----------
tangle <- read_csv("tanglegramdogs.csv", show_col_types = FALSE) %>%
  mutate(
    Assigned_haplogroup = na_if(trimws(as.character(Assigned_haplogroup)), ""),
    Beast_node_defining_clade = suppressWarnings(as.integer(Beast_node_defining_clade)),
    ML_node_defining_clade = suppressWarnings(as.integer(ML_node_defining_clade))
  )

# Keep only rows with a valid BEAST clade node
clade_map <- tangle %>%
  filter(!is.na(Beast_node_defining_clade))

# -------- load ML tree ----------
ml_treed <- read.tree("dogs_it7_ML.tree")
ml_treed$tip.label <- gsub("_extraction$", "", ml_treed$tip.label)


# -------- compute BEAST vs ML similarity ----------
support_prop <- numeric(nrow(clade_map))

for (i in seq_len(nrow(clade_map))) {
  beast_node <- clade_map$Beast_node_defining_clade[i]
  ml_node <- clade_map$ML_node_defining_clade[i]
  
  beast_desc <- get_descendant_labels(tree, beast_node)
  
  if (is.na(ml_node)) {
    prop <- 0
  } else {
    ml_desc <- get_descendant_labels(ml_treed, ml_node)
    overlap <- sum(beast_desc %in% ml_desc)
    prop <- overlap / length(beast_desc)
  }
  
  support_prop[i] <- prop
}

clade_map$support_prop <- support_prop
clade_map$tier <- case_when(
  support_prop >= 0.95 ~ ">=0.95",
  support_prop >= 0.50 ~ "0.50-0.95",
  TRUE ~ "<0.50"
)
clade_map$tier <- factor(clade_map$tier, levels = c("<0.50", "0.50-0.95", ">=0.95"))


# -------- build ggtree object ----------
p5 <- ggtree(beast_tree, layout = "rectangular", branch.length = "none")

# Ensure plotting data has labels
if (!"label" %in% names(p5$data) || all(is.na(p5$data$label))) {
  p5$data$label <- beast_tree@data$label
} else {
  missing_labels <- is.na(p5$data$label) & !is.null(beast_tree@data$label)
  if (any(missing_labels)) p5$data$label[missing_labels] <- beast_tree@data$label[missing_labels]
}

# Ensure posterior column exists
if (!"posterior" %in% names(p5$data)) p5$data$posterior <- NA
if (!is.null(beast_tree@data$posterior)) {
  missing_post <- is.na(p5$data$posterior) & !is.na(beast_tree@data$posterior)
  if (any(missing_post)) p5$data$posterior[missing_post] <- beast_tree@data$posterior[missing_post]
}

# Attach clade colours
p5$data$clade_colour <- node_colours[p5$data$node]
p5$data$clade_colour[is.na(p5$data$clade_colour)] <- "#000000"

# Major clade-defining nodes from tanglegramdogs.csv
major_nodes <- clade_map$Beast_node_defining_clade

# Label the major nodes, plus all other internal nodes with posterior >= 0.5
p5$data$show_major <- !p5$data$isTip & p5$data$node %in% major_nodes & !is.na(p5$data$posterior)
p5$data$show_minor <- !p5$data$isTip & p5$data$posterior >= 0.5 & !(p5$data$node %in% major_nodes)

p5$data$support_label <- ifelse(
  !is.na(p5$data$posterior),
  sprintf("%.2f", p5$data$posterior),
  ""
)

# Posterior text colour should match clade colour
p5$data$support_colour <- p5$data$clade_colour

# Nodes to draw circles on: only BEAST clade-defining nodes from tanglegramdogs.csv
internal_nodes_plot <- p5$data %>%
  filter(node %in% major_nodes) %>%
  left_join(
    clade_map %>% select(Beast_node_defining_clade, support_prop, tier),
    by = c("node" = "Beast_node_defining_clade")
  )

tier_cols <- c(
  ">=0.95" = "#08306B",
  "0.50-0.95" = "#6BAED6",
  "<0.50" = "white"
)

# -------- plot ----------
p5.5 <- p5 +
  geom_tree(aes(colour = clade_colour)) +
  geom_tiplab(
    aes(label = label, colour = clade_colour),
    size = 3,
    align = TRUE,
    linetype = "dotted",
    show.legend = FALSE
  ) +
  geom_text2(
    aes(
      subset = show_major,
      label = support_label,
      colour = support_colour
    ),
    size = 3.2,
    fontface = "bold",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_text2(
    aes(
      subset = show_minor,
      label = support_label,
      colour = support_colour
    ),
    size = 2,
    fontface = "plain",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_point(
    data = internal_nodes_plot,
    aes(x = x, y = y, fill = tier),
    shape = 21,
    size = 3,
    stroke = 0.3,
    colour = "black"
  ) +
  scale_fill_manual(values = tier_cols, name = "BEAST vs ML similarity", drop = FALSE) +
  scale_colour_identity() +
  coord_cartesian(clip = "off") +
  theme_tree() +
  theme(
    plot.margin = margin(20, 60, 20, 40)
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9))) +
  theme(legend.position = "none")

p5.5
```

## Humans

```{r}
#| results: 'hide'
#| fig-width: 15
#| fig-height: 60
#| fig-cap: "Tree comparing human ML and BEAST tree topology. Foundation is the BEAST tree with circles on major nodes correspondign to ML agreement of taxa inside major clades (dark blue = 95% similarity, light blue = > 50% similarity, white = <50% similarity)."
#| fig-align: center

library(treeio)
library(ggtree)
library(ggplot2)
library(ape)
library(readr)
library(dplyr)

options(ignore.negative.edge = TRUE)

relaxed_thresh <- 0.95

# -------- helper functions ----------
get_descendant_tips <- function(phy, node) {
  children <- phy$edge[phy$edge[, 1] == node, 2]
  tips <- integer(0)
  for (ch in children) {
    if (ch <= length(phy$tip.label)) {
      tips <- c(tips, ch)
    } else {
      tips <- c(tips, get_descendant_tips(phy, ch))
    }
  }
  unique(tips)
}

get_descendant_labels <- function(phy, node) {
  idx <- get_descendant_tips(phy, node)
  phy$tip.label[idx]
}

getDescendants <- function(tree, node) {
  children <- tree$edge[tree$edge[, 1] == node, 2]
  all_desc <- children
  for (child in children) {
    all_desc <- c(all_desc, getDescendants(tree, child))
  }
  return(all_desc)
}
# ------------------------------------

# -------- load BEAST tree ----------
beast_tree <- read.beast("humans_it14_517_allruns_MCC_rooted.tree")
tree <- as.phylo(beast_tree)

# Clean tip labels if needed
tree$tip.label <- gsub("_extraction$", "", tree$tip.label)
beast_tree@phylo$tip.label <- tree$tip.label


# -------- load ML tree ----------
ml_tree <- read.tree("humans_it14_517_aligned_inspected_ML.tree")
ml_tree$tip.label <- gsub("_extraction$", "", ml_tree$tip.label)
ml_tree <- root(ml_tree, outgroup = "AF347014.1_yoruba_reference", resolve.root = TRUE)

# -------- read human BEAST colouring table ----------
clade_defs <- read_csv("human_it14_beast_tree_colour.csv", show_col_types = FALSE) %>%
  mutate(
    Fasta_ID_alignment = as.character(Fasta_ID_alignment),
    Assigned_Hg_based_on_tree = na_if(trimws(as.character(Assigned_Hg_based_on_tree)), ""),
    node_of_assigned_monophyletic_clade = suppressWarnings(as.integer(node_of_assigned_monophyletic_clade)),
    colour_name = na_if(trimws(as.character(colour_name)), ""),
    colour_hex_code = na_if(trimws(as.character(colour_hex_code)), "")
  )

clade_rows <- clade_defs %>%
  filter(
    !is.na(Assigned_Hg_based_on_tree),
    !is.na(node_of_assigned_monophyletic_clade),
    !is.na(colour_hex_code)
  )

# Default all nodes/tips to black
node_colours <- rep("#000000", Ntip(tree) + Nnode(tree))

# Assign clade colours to each defined node + all descendants
for (i in seq_len(nrow(clade_rows))) {
  node_id <- clade_rows$node_of_assigned_monophyletic_clade[i]
  hex <- clade_rows$colour_hex_code[i]
  desc <- getDescendants(tree, node_id)
  
  node_colours[node_id] <- hex
  if (length(desc) > 0) node_colours[desc] <- hex
}

# -------- read tanglegram table ----------
tangle <- read_csv("tanglegram.csv", show_col_types = FALSE) %>%
  mutate(
    Assigned_haplogroup = na_if(trimws(as.character(Assigned_haplogroup)), ""),
    Beast_node_defining_clade = suppressWarnings(as.integer(Beast_node_defining_clade)),
    ML_node_defining_clade = suppressWarnings(as.integer(ML_node_defining_clade))
  )

# Keep only rows with a valid BEAST clade node
clade_map <- tangle %>%
  filter(!is.na(Beast_node_defining_clade))

# -------- compute BEAST vs ML similarity ----------
support_prop <- numeric(nrow(clade_map))

for (i in seq_len(nrow(clade_map))) {
  beast_node <- clade_map$Beast_node_defining_clade[i]
  ml_node <- clade_map$ML_node_defining_clade[i]
  
  beast_desc <- get_descendant_labels(tree, beast_node)
  
  if (is.na(ml_node)) {
    prop <- 0
  } else {
    ml_desc <- get_descendant_labels(ml_tree, ml_node)
    overlap <- sum(beast_desc %in% ml_desc)
    prop <- overlap / length(beast_desc)
  }
  
  support_prop[i] <- prop
}

clade_map$support_prop <- support_prop
clade_map$tier <- case_when(
  support_prop >= 0.95 ~ ">=0.95",
  support_prop >= 0.50 ~ "0.50-0.95",
  TRUE ~ "<0.50"
)
clade_map$tier <- factor(clade_map$tier, levels = c("<0.50", "0.50-0.95", ">=0.95"))

# -------- build ggtree object ----------
p6 <- ggtree(beast_tree, layout = "rectangular", branch.length = "none")

# Ensure plotting data has labels
if (!"label" %in% names(p6$data) || all(is.na(p6$data$label))) {
  p6$data$label <- beast_tree@data$label
} else {
  missing_labels <- is.na(p6$data$label) & !is.null(beast_tree@data$label)
  if (any(missing_labels)) p6$data$label[missing_labels] <- beast_tree@data$label[missing_labels]
}

# Ensure posterior column exists
if (!"posterior" %in% names(p6$data)) p6$data$posterior <- NA
if (!is.null(beast_tree@data$posterior)) {
  missing_post <- is.na(p6$data$posterior) & !is.na(beast_tree@data$posterior)
  if (any(missing_post)) p6$data$posterior[missing_post] <- beast_tree@data$posterior[missing_post]
}

# Attach clade colours
p6$data$clade_colour <- node_colours[p6$data$node]
p6$data$clade_colour[is.na(p6$data$clade_colour)] <- "#000000"

# Major clade-defining nodes from tanglegram.csv
major_nodes <- clade_map$Beast_node_defining_clade

# Label the major nodes, plus all other internal nodes with posterior >= 0.5
p6$data$show_major <- !p6$data$isTip & p6$data$node %in% major_nodes & !is.na(p6$data$posterior)
p6$data$show_minor <- !p6$data$isTip & p6$data$posterior >= 0.5 & !(p6$data$node %in% major_nodes)

p6$data$support_label <- ifelse(
  !is.na(p6$data$posterior),
  sprintf("%.2f", p6$data$posterior),
  ""
)

# Posterior text colour should match clade colour
p6$data$support_colour <- p6$data$clade_colour

# Nodes to draw circles on: only BEAST clade-defining nodes from tanglegram.csv
internal_nodes_plot <- p6$data %>%
  filter(node %in% major_nodes) %>%
  left_join(
    clade_map %>% select(Beast_node_defining_clade, support_prop, tier),
    by = c("node" = "Beast_node_defining_clade")
  )

tier_cols <- c(
  ">=0.95" = "#08306B",
  "0.50-0.95" = "#6BAED6",
  "<0.50" = "white"
)

# -------- plot ----------
p6.6 <- p6 +
  geom_tree(aes(colour = clade_colour)) +
  geom_tiplab(
    aes(label = label, colour = clade_colour),
    size = 3,
    align = TRUE,
    linetype = "dotted",
    show.legend = FALSE
  ) +
  geom_text2(
    aes(
      subset = show_major,
      label = support_label,
      colour = support_colour
    ),
    size = 3.2,
    fontface = "bold",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_text2(
    aes(
      subset = show_minor,
      label = support_label,
      colour = support_colour
    ),
    size = 2,
    fontface = "plain",
    hjust = -0.3,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_point(
    data = internal_nodes_plot,
    aes(x = x, y = y, fill = tier),
    shape = 21,
    size = 3,
    stroke = 0.3,
    colour = "black"
  ) +
  scale_fill_manual(values = tier_cols, name = "BEAST vs ML similarity", drop = FALSE) +
  scale_colour_identity() +
  coord_cartesian(clip = "off") +
  theme_tree() +
  theme(
    plot.margin = margin(20, 60, 20, 40)
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.9))) + 
  theme(legend.position = "none")

p6.6
```
